A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-07 paper.tex

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As we demonstrate in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
might seem of little practical importance.
However, as we show in the remainder of this article,
there are substantial practical gains
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computationally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.

@@

In practice, the backwards-in-time Griffiths process that produces
ARGs such as that shown in Figs~{fig-ancestry-resolution}A~and~B
is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
dominated by nodes such as \textsf{g} and \textsf{j} which contain information
about inheritance in non-ancestral material, as well as nodes such as
\textsf{o} and \textsf{q} which track details of ancestry even once the
local genomic region has fully coalesced. Instead, simulating
ARGs from the coalescent with recombination is usually done by
running Hudson's algorithm to produce a sample-resolved ARG directly.
Moreover, when inferring ARGs from genetic data, there is no information
in the sample set that can be used to resolve these nodes, so that they
are essentially unknowable.

@@

It is worth noting the process of sample resolution can reveal that certain ARG
nodes are \emph{redundant}: if they are removed and the edges adjusted to
pas directly from the node's child to the node's parent, it does not reduce our
knowledge of the samples' genetic ancestry. Removing nodes,
both redundant ones and others, is part of a process we call
``simplification''.

@@

Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
explicit recombination events we cannot compute a likelihood
under the coalescent with recombination (see appendix XXX).
[Some stuff about whether we should see this as necessary or not,
but maybe we should keep this for the conclusion?]

>>>>

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As demonstrated in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
might seem of little practical importance.
However, there are substantial practical benefits
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computationally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.

@@

In practice, the backwards-in-time Griffiths process that produces
ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
dominated by nodes such as \textsf{g} and \textsf{j} which contain information
about inheritance in non-ancestral material, as well as nodes such as
\textsf{o} and \textsf{q} which track details of ancestry even once the
local genomic region has fully coalesced. Instead, simulating
ARGs from the coalescent with recombination is usually done by
running Hudson's algorithm to produce a sample-resolved ARG directly.
Moreover, when inferring ARGs from genetic data, there is no information
in the sample set that can be used to resolve these nodes, so that they
are essentially unknowable.

@@

% It is worth noting the process of sample resolution can reveal that certain ARG
% nodes are \emph{redundant}: if they are removed and the edges adjusted to
% pas directly from the node's child to the node's parent, it does not reduce our
% knowledge of the samples' genetic ancestry. Removing nodes,
% both redundant ones and others, is part of a process we call
% ``simplification''.

@@

% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
% explicit recombination events we cannot compute a likelihood
% under the coalescent with recombination (see appendix XXX).
% [Some stuff about whether we should see this as necessary or not,
% but maybe we should keep this for the conclusion?]
